---
title: "mmLung_Spleen_bulkRNA"
author: "Jaehyun Choi"
date: "2025-09-02"
output: html_document
---

```{r setup, include=FALSE}
library(Seurat)
library(MAST)
library(dittoSeq)
library(caret)
library(ranger)
library(feseR)
#  install.packages("devtools")
library(devtools)
#install_github("enriquea/feseR")
library(feseR)
library(pROC)
library(reticulate)
library(SingleR)
library(slingshot)
library(celldex)
library(scCATCH)
library(dplyr)
```

# Count importing
```{r}
setwd("~/bystander_collaboration/bulk_data/count")

samples <- c(
"sample1_count",
"sample2_count",
"sample3_count",
"sample4_count",
"sample5_count",
"sample6_count",
"sample7_count",
"sample8_count",
"sample9_count",
"sample10_count",
"sample11_count",
"sample12_count"
);
infiles <- paste(samples, ".out", sep="");
inlist <- lapply(infiles, read.table);
alltrans <- unique(unlist(lapply(inlist, "[[", 1), use.names=FALSE));

## construct empty count matrix
CM <- structure(matrix(as.integer(NA),
                       nrow=length(alltrans),
                       ncol=length(samples)),
                dimnames=list(alltrans, samples))

## fill count matrix
for (i in 1:length(samples)) {
  CM[match(inlist[[i]]$V1, rownames(CM)), i] <- inlist[[i]]$V2;
}

## drop non-unique alignments and other non-gene entries
dropClasses <- c("__alignment_not_unique", "__no_feature", "__ambiguous",
                 "__too_low_aQual", "__not_aligned")

if (any(dropClasses %in% rownames(CM)))
  CM <- CM[-match(dropClasses, rownames(CM)),]

colnames(CM) <- c(
"mmLung_highPD1_MPCD4T_Rep1",
"mmLung_lowPD1_MPCD4T_Rep1", 
"mmLung_highPD1_MPCD4T_Rep2", 
"mmLung_lowPD1_MPCD4T_Rep2",
"mmLung_highPD1_MPCD4T_Rep3",
"mmLung_lowPD1_MPCD4T_Rep3",
"mmSpleen_highPD1_MPCD4T_Rep1",
"mmSpleen_lowPD1_MPCD4T_Rep1",
"mmSpleen_highPD1_MPCD4T_Rep2",
"mmSpleen_lowPD1_MPCD4T_Rep2",
"mmSpleen_highPD1_MPCD4T_Rep3",
"mmSpleen_lowPD1_MPCD4T_Rep3"
)
#write.csv(as.data.frame(CM), file="CM.csv")
```

## bulk data QC

```{r}
setwd("~/bystander_collaboration/bulk_data/analysis")

barplot(colSums(CM), col="sienna", las=2, cex.names =0.6, cex.axis=0.6, ylab = "library size")
mtext(side=2, line=4.5, text="Number of transcript-counted reads", font=2)

#pdf("bulkRNA_barplot_IPF.pdf", width = 6, height = 5)
#p <- barplot(colSums(CM), col="sienna", las=2, cex.names =0.6, cex.axis=0.6, ylab = "library size")
#mtext(side=2, line=4.5, text="Number of transcript-counted reads", font=2)
#print(p)
#dev.off()

## Make condition column for the sample information
cts <- CM
coldata <- as.data.frame(c("Lung_PD1high","Lung_PD1low","Lung_PD1high","Lung_PD1low","Lung_PD1high","Lung_PD1low","Spleen_PD1high","Spleen_PD1low","Spleen_PD1high","Spleen_PD1low","Spleen_PD1high","Spleen_PD1low"))
rownames(coldata) <- colnames(cts)
colnames(coldata) <- "condition"

## To extract normalized read count
library("DESeq2")
dds <- DESeqDataSetFromMatrix(countData = cts, colData = coldata, design = ~ condition)
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
dds <- DESeq(dds)
normcount <- counts(dds, normalized=TRUE)
#write.csv(as.data.frame(normcount), file="normcount.csv")

## HVG analysis 
gene_vars <- apply(normcount, 1, var)
top3000_genes <- names(sort(gene_vars, decreasing = TRUE))[1:3000]
normcount_top3000 <- normcount[top3000_genes, ]

### correlation:
CM.cor <- cor(normcount, method="spearman")
## general correlation between samples:
CM.dists <- 1-as.dist(cor(normcount, method="spearman"))
CM.hc <- hclust(CM.dists)
plot(CM.hc, main=NA, xlab="Spearman correlation distance")

## general correlation between samples:  
CM.dists <- 1 - as.dist(cor(normcount_top3000, method = "spearman"))
CM.hc <- hclust(CM.dists)  
plot(CM.hc, main = "Sample Clustering based on HVG 3000", xlab = "Spearman correlation distance")
```

## PCA analysis
```{r}
# lung 샘플 (앞의 6개)
lung_samples <- c(
  "mmLung_highPD1_MPCD4T_Rep1",
  "mmLung_lowPD1_MPCD4T_Rep1", 
  "mmLung_highPD1_MPCD4T_Rep2", 
  "mmLung_lowPD1_MPCD4T_Rep2",
  "mmLung_highPD1_MPCD4T_Rep3",
  "mmLung_lowPD1_MPCD4T_Rep3"
)

# spleen 샘플 (7번째부터 끝까지, sample7은 주석처리되어 제외됨)
spleen_samples <- c(
  "mmSpleen_highPD1_MPCD4T_Rep1",
  "mmSpleen_lowPD1_MPCD4T_Rep1",
  "mmSpleen_highPD1_MPCD4T_Rep2",
  "mmSpleen_lowPD1_MPCD4T_Rep2",
  "mmSpleen_highPD1_MPCD4T_Rep3",
  "mmSpleen_lowPD1_MPCD4T_Rep3"
)

# 분리된 matrix
lung_norm <- normcount[, lung_samples]
spleen_norm <- normcount[, spleen_samples]

## lung PCA
lung_norm_log2 <- log2(lung_norm + 1)
lung_var <- apply(lung_norm_log2, 1, sd)
lung_norm_filtered <- lung_norm_log2[lung_var > 0, ]

data_sub_t <- as.data.frame(t(lung_norm_filtered))
pca_result <- prcomp(data_sub_t, center = TRUE, scale. = TRUE)
ind.coord <- pca_result$x

lung_colors <- c("red", "blue", "red", "blue","red", "blue")
lung_colors <- c("red", "blue", "orange", "skyblue", "pink", "green")

plot(ind.coord[,1], ind.coord[,2], col=lung_colors, pch=16, xlab="PC1", ylab="PC2")
legend("topright", legend=c("PD1-high", "PD1-low"), col=c("red", "blue"), pch=16)

plot(ind.coord[,1], ind.coord[,2], col = lung_colors, pch = 16, xlab = "PC1", ylab = "PC2")
text(ind.coord[,1], ind.coord[,2], labels = rownames(ind.coord), pos = 3, cex = 0.7)

lung_colors <- c("red", "blue", "red", "blue","red", "blue")
pdf("bulkRNA_lungIPF_PCA.pdf", width = 6, height = 5)
lung_colors <- c("red", "blue", "red", "blue","red", "blue")
# 플롯 생성
p <- plot(ind.coord[,1], ind.coord[,2], col = lung_colors, pch = 16, xlab = "PC1", ylab = "PC2")
#text(ind.coord[,1], ind.coord[,2], labels = rownames(ind.coord), pos = 3, cex = 0.7)

# 플롯 출력
print(p)
dev.off()

lung_colors <- c("red", "blue", "orange", "skyblue", "pink", "green")
pdf("bulkRNA_lungIPF_PCA_specific.pdf", width = 6, height = 5)

# 플롯 생성
p <- plot(ind.coord[,1], ind.coord[,2], col = lung_colors, pch = 16, xlab = "PC1", ylab = "PC2")
text(ind.coord[,1], ind.coord[,2], labels = rownames(ind.coord), pos = 3, cex = 0.7)

# 플롯 출력
print(p)
dev.off()


# Lung Eigenvalues
eig <- (pca_result$sdev)^2
# Variances in percentage
variance <- eig*100/sum(eig)
# Cumulative variances
cumvar <- cumsum(variance)
eig.decathlon2.active <- data.frame(eig = eig, variance = variance,
                     cumvariance = cumvar)
head(eig.decathlon2.active)


#barplot(eig.decathlon2.active[, 2], names.arg=1:nrow(eig.decathlon2.active), 
barplot(eig.decathlon2.active[1:10, 2], names.arg=1:10, 
       main = "Variances",
       xlab = "Principal Components",
       ylab = "Percentage of variances",
       col ="steelblue")
# Add connected line segments to the plot
lines(x = 1:10, 
      eig.decathlon2.active[1:10, 2], 
      type="b", pch=19, col = "red")

pdf("eigen_bargraph_IPF.pdf", width = 6, height = 5)

# 플롯 생성
p <- barplot(eig.decathlon2.active[1:10, 2], names.arg=1:10, 
       main = "Variances",
       xlab = "Principal Components",
       ylab = "Percentage of variances",
       col ="steelblue")
# Add connected line segments to the plot
lines(x = 1:10, 
      eig.decathlon2.active[1:10, 2], 
      type="b", pch=19, col = "red")

# 플롯 출력
print(p)
dev.off()


lung_log2 <- log2(lung_norm + 1)
pairs(lung_log2,
      main = "Pairwise Scatter Plot: Lung Samples",
      pch = 19, cex = 0.5)

library(GGally)

lung_log2_t <- as.data.frame(t(lung_log2))
lung_log2_t$Group <- factor(c("PD1-high", "PD1-low", "PD1-high", "PD1-low", "PD1-high", "PD1-low"))

# ggpairs 그리기
GGally::ggpairs(lung_log2_t[, 1:(ncol(lung_log2_t)-1)],   # 유전자들만
                title = "Pairwise Scatter (Lung samples)",
                mapping = ggplot2::aes(color = lung_log2_t$Group),
                lower = list(continuous = "points"),
                diag = list(continuous = "densityDiag"),
                upper = list(continuous = "cor"))


## Spleen PCA
spleen_norm_log2 <- log2(spleen_norm + 1)
spleen_var <- apply(spleen_norm_log2, 1, sd)
spleen_norm_filtered <- spleen_norm_log2[spleen_var > 0, ]

data_sub_t <- as.data.frame(t(spleen_norm_filtered))
pca_result <- prcomp(data_sub_t, center = TRUE, scale. = TRUE)
ind.coord <- pca_result$x

spleen_colors <- c("red", "blue", "red", "blue", "red", "blue")
spleen_colors <- c("red", "blue", "orange", "skyblue", "pink", "green")

plot(ind.coord[,1], ind.coord[,2], col=spleen_colors, pch=16, xlab="PC1", ylab="PC2")
legend("topright", legend=c("PD1-high", "PD1-low"), col=c("red", "blue"), pch=16)

plot(ind.coord[,1], ind.coord[,2], col = spleen_colors, pch = 16, xlab = "PC1", ylab = "PC2")
text(ind.coord[,1], ind.coord[,2], labels = rownames(ind.coord), pos = 3, cex = 0.7)

spleen_colors <- c("red", "blue", "red", "blue","red", "blue")
pdf("bulkRNA_spleenIPF_PCA.pdf", width = 6, height = 5)
spleen_colors <- c("red", "blue", "red", "blue","red", "blue")
# 플롯 생성
p <- plot(ind.coord[,1], ind.coord[,2], col = spleen_colors, pch = 16, xlab = "PC1", ylab = "PC2")
#text(ind.coord[,1], ind.coord[,2], labels = rownames(ind.coord), pos = 3, cex = 0.7)

# 플롯 출력
print(p)
dev.off()

spleen_colors <- c("red", "blue", "orange", "skyblue", "pink", "green")
pdf("bulkRNA_spleenIPF_PCA_specific.pdf", width = 6, height = 5)

# 플롯 생성
p <- plot(ind.coord[,1], ind.coord[,2], col = spleen_colors, pch = 16, xlab = "PC1", ylab = "PC2")
text(ind.coord[,1], ind.coord[,2], labels = rownames(ind.coord), pos = 3, cex = 0.7)

# 플롯 출력
print(p)
dev.off()



# Lung Eigenvalues
eig <- (pca_result$sdev)^2
# Variances in percentage
variance <- eig*100/sum(eig)
# Cumulative variances
cumvar <- cumsum(variance)
eig.decathlon2.active <- data.frame(eig = eig, variance = variance,
                     cumvariance = cumvar)
head(eig.decathlon2.active)


#barplot(eig.decathlon2.active[, 2], names.arg=1:nrow(eig.decathlon2.active), 
barplot(eig.decathlon2.active[1:10, 2], names.arg=1:10, 
       main = "Variances",
       xlab = "Principal Components",
       ylab = "Percentage of variances",
       col ="steelblue")
# Add connected line segments to the plot
lines(x = 1:10, 
      eig.decathlon2.active[1:10, 2], 
      type="b", pch=19, col = "red") 
```

## Lung analysis

```{r}

library(org.Mm.eg.db)
library(clusterProfiler)

##  subset to lung only
lung_samples <- grep("^mmLung", colnames(CM), value = TRUE)
lung_counts <- CM[, lung_samples]

## Make condition column for the sample information
cts <- lung_counts
coldata <- as.data.frame(c("Lung_PD1high","Lung_PD1low","Lung_PD1high","Lung_PD1low","Lung_PD1high","Lung_PD1low"))
rownames(coldata) <- colnames(cts)
colnames(coldata) <- "condition"

# DESeq2 object
dds <- DESeqDataSetFromMatrix(countData = cts, colData = coldata, design = ~ condition)
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
dds <- DESeq(dds)

res <- results(dds, contrast=c("condition", "Lung_PD1high","Lung_PD1low"))

plotMA(res, ylim=c(-2,2))
plotCounts(dds, gene=which.min(res$padj), intgroup="condition")

# Heatmap
res$DEG <- 2

res[res$padj<0.05 & res$log2FoldChange>0.58 & is.na(res$padj)==0, 7] <- 1 ##Up-regulated genes
res[res$padj<0.05 & res$log2FoldChange<(-0.58) & is.na(res$padj)==0, 7] <- 3 ##Down-regulated genes
library (pheatmap)
library (gplots)

breaksList = seq(-3, 3, by = 0.1)
breaksList=append(breaksList, 20)
breaksList=append(breaksList, -20, 0)
mycol <- colorpanel(n=length(breaksList)-1,low="blue", mid = "white", high="red")

mat <- normcount[res$DEG != 2, 1:6]
mat <- mat[complete.cases(mat), ]
mat <- mat[apply(mat, 1, sd) != 0, ]
mat <- mat[!rowSums(!is.finite(as.matrix(mat))), ]

pheatmap(mat,
         show_rownames = FALSE,
         show_colnames = TRUE,
         cluster_cols = FALSE,
         cluster_rows = TRUE,
         scale = "row",
         col = mycol,
         breaks = breaksList,
         clustering_distance_cols = "euclidean",
         cex = 0.8)

# DEG 유전자 중에서 log2FC 기준 정렬
selected_genes <- rownames(res)[res$DEG != 2]
res_sub <- res[selected_genes, ]
res_sub <- res_sub[order(res_sub$log2FoldChange, decreasing = TRUE), ]

# 정렬된 유전자 순서로 count matrix 추출
mat <- normcount[rownames(res_sub), 1:6]

# 필터링 (NA, sd=0, Inf 등 제거)
mat <- mat[complete.cases(mat), ]
mat <- mat[apply(mat, 1, sd) != 0, ]
mat <- mat[!rowSums(!is.finite(as.matrix(mat))), ]

# Heatmap (cluster_rows = FALSE 하면 정렬 순서 유지)
pheatmap(mat,
         show_rownames = FALSE,
         show_colnames = TRUE,
         cluster_cols = TRUE,
         cluster_rows = FALSE,  # ✅ 정렬된 순서대로 보여주기 위해 FALSE
         scale = "row",
         col = mycol,
         breaks = breaksList,
         clustering_distance_cols = "euclidean",
         cex = 0.8)

```
pdf("bulkRNA_heatmap_mmLung.pdf", width = 6, height = 5)

pheatmap(mat,
         show_rownames = FALSE,
         show_colnames = TRUE,
         cluster_cols = FALSE,
         cluster_rows = FALSE,
         scale = "row",
         col = mycol,
         breaks = breaksList,
         clustering_distance_cols = "euclidean",
         cex = 0.8)

dev.off()

``` {r}
# ENSEMBL to Gene Symbol
summary(res)

rownames(res) <- sub("\\..*", "", rownames(res))

# ENSEMBL → SYMBOL 매핑
gene_df <- bitr(rownames(res),
                fromType = "ENSEMBL",
                toType = "SYMBOL",
                OrgDb = org.Mm.eg.db)
                
# DESeq2 결과(res)에 SYMBOL 붙이기
res$SYMBOL <- gene_df$SYMBOL[match(rownames(res), gene_df$ENSEMBL)]

write.csv(as.data.frame(res), file="DESeq_result_mmLung.csv")


```

## Spleen analysis

```{r}

##  subset to Spleen only
spleen_samples <- grep("^mmSpleen", colnames(CM), value = TRUE)
spleen_counts <- CM[, spleen_samples]

## Make condition column for the sample information
cts <- spleen_counts
coldata <- as.data.frame(c("Spleen_PD1high","Spleen_PD1low","Spleen_PD1high","Spleen_PD1low","Spleen_PD1high","Spleen_PD1low"))
rownames(coldata) <- colnames(cts)
colnames(coldata) <- "condition"

# DESeq2 object
dds <- DESeqDataSetFromMatrix(countData = cts, colData = coldata, design = ~ condition)
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
dds <- DESeq(dds)

res <- results(dds, contrast=c("condition", "Spleen_PD1high","Spleen_PD1low"))

plotMA(res, ylim=c(-2,2))
plotCounts(dds, gene=which.min(res$padj), intgroup="condition")
# Heatmap
res$DEG <- 2

res[res$padj<0.05 & res$log2FoldChange>0.58 & is.na(res$padj)==0, 7] <- 1 ##Up-regulated genes
res[res$padj<0.05 & res$log2FoldChange<(-0.58) & is.na(res$padj)==0, 7] <- 3 ##Down-regulated genes
library (pheatmap)
library (gplots)

breaksList = seq(-3, 3, by = 0.1)
breaksList=append(breaksList, 20)
breaksList=append(breaksList, -20, 0)
mycol <- colorpanel(n=length(breaksList)-1,low="blue", mid = "white", high="red")

mat <- normcount[res$DEG != 2, 1:6]
mat <- mat[complete.cases(mat), ]
mat <- mat[apply(mat, 1, sd) != 0, ]
mat <- mat[!rowSums(!is.finite(as.matrix(mat))), ]

pheatmap(mat,
         show_rownames = FALSE,
         show_colnames = TRUE,
         cluster_cols = FALSE,
         cluster_rows = TRUE,
         scale = "row",
         col = mycol,
         breaks = breaksList,
         clustering_distance_cols = "euclidean",
         cex = 0.8)

# DEG 유전자 중에서 log2FC 기준 정렬
selected_genes <- rownames(res)[res$DEG != 2]
res_sub <- res[selected_genes, ]
res_sub <- res_sub[order(res_sub$log2FoldChange, decreasing = TRUE), ]

# 정렬된 유전자 순서로 count matrix 추출
mat <- normcount[rownames(res_sub), 1:6]

# 필터링 (NA, sd=0, Inf 등 제거)
mat <- mat[complete.cases(mat), ]
mat <- mat[apply(mat, 1, sd) != 0, ]
mat <- mat[!rowSums(!is.finite(as.matrix(mat))), ]

# Heatmap (cluster_rows = FALSE 하면 정렬 순서 유지)
pheatmap(mat,
         show_rownames = FALSE,
         show_colnames = TRUE,
         cluster_cols = TRUE,
         cluster_rows = FALSE,  # ✅ 정렬된 순서대로 보여주기 위해 FALSE
         scale = "row",
         col = mycol,
         breaks = breaksList,
         clustering_distance_cols = "euclidean",
         cex = 0.8)

```
pdf("bulkRNA_heatmap_mmLung.pdf", width = 6, height = 5)

pheatmap(mat,
         show_rownames = FALSE,
         show_colnames = TRUE,
         cluster_cols = FALSE,
         cluster_rows = FALSE,
         scale = "row",
         col = mycol,
         breaks = breaksList,
         clustering_distance_cols = "euclidean",
         cex = 0.8)

dev.off()
``` {r}
# ENSEMBL to Gene Symbol
summary(res)

rownames(res) <- sub("\\..*", "", rownames(res))

# ENSEMBL → SYMBOL 매핑
gene_df <- bitr(rownames(res),
                fromType = "ENSEMBL",
                toType = "SYMBOL",
                OrgDb = org.Mm.eg.db)
                
# DESeq2 결과(res)에 SYMBOL 붙이기
res$SYMBOL <- gene_df$SYMBOL[match(rownames(res), gene_df$ENSEMBL)]

write.csv(as.data.frame(res), file="DESeq_result_mmSpleen.csv")


```
