---
title: "Byst_monocle3"
output: html_document
date: "2025-04-03"
---

```{r setup, include=FALSE}
library(Seurat)
library(SeuratWrappers)
library(monocle3)
```


```{r cars}
Cord_byst <- readRDS("/home/jaehyunchoi/project/Byst_collaboration/analysis/Cord_bystR.rds")
cds <- as.cell_data_set(Cord_byst, )
cds

cds <- reduce_dimension(cds)
cds <- reduce_dimension(cds, reduction_method = "UMAP")
cds <- cluster_cells(cds)
cds <- learn_graph(cds)
colData(cds)$Celltype <- Cord_byst$celltype
root_cells <- colnames(cds)[cds@colData$Celltype == "PD-1_high"]
plot_cells(cds, color_cells_by = "Celltype")

pdf("/home/jaehyunchoi/project/Byst_collaboration/analysis/monocle3_plot_1.pdf", width = 8, height = 6)
plot_cells(cds, color_cells_by = "Celltype")
dev.off()
pdf("/home/jaehyunchoi/project/Byst_collaboration/analysis/monocle3_plot_2.pdf", width = 8, height = 6)
plot_cells(cds, label_groups_by_cluster=FALSE,  color_cells_by = "Celltype")
dev.off()
pdf("/home/jaehyunchoi/project/Byst_collaboration/analysis/monocle3_plot_3.pdf", width = 8, height = 6)
plot_cells(cds,
           color_cells_by = "pseudotime",
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points=FALSE,
           graph_label_size=1.5)
dev.off()
cds <- order_cells(cds, root_cells = root_cells[1])
cds <- order_cells(cds)
pseudotime(cds)


pdf("/home/jaehyunchoi/project/Byst_collaboration/analysis/monocle3_plot.pdf", width = 8, height = 6)
plot_cells(cds, label_groups_by_cluster = FALSE, color_cells_by = "celltype")
dev.off()
```

# Harmony 진행한 이후

```{r}
Cord_bysT <- readRDS("/home/jaehyunchoi/project/Byst_collaboration/analysis/Cord_byst_ver2.rds")
cds1 <- as.cell_data_set(Cord_bysT, )
cds1
colData(cds1)$Celltype <- Cord_bysT$celltype
pdf("/home/jaehyunchoi/project/Byst_collaboration/analysis/monocle3_plot_hm1.pdf", width = 8, height = 6)
plot_cells(cds1, color_cells_by = "Celltype")
dev.off()
#cds1 <- reduce_dimension(cds1, reduction_method = "UMAP")
cds1 <- cluster_cells(cds1)
cds1 <- learn_graph(cds1)
cds1 <- order_cells(cds1)
pseudotime(cds1)

pdf("/home/jaehyunchoi/project/Byst_collaboration/analysis/monocle3_plot_1_2.pdf", width = 8, height = 6)
plot_cells(cds1,
           color_cells_by = "pseudotime",
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points=FALSE,
           graph_label_size=1.5)
dev.off()

cds1_sub <- choose_graph_segments(cds1)


colData(cds1)$Celltype <- Cord_byst$celltype
root_cells <- colnames(cds1)[cds1@colData$Celltype == "PD-1_high"]
plot_cells(cds1, color_cells_by = "Celltype")
```

```{r cars}
# preprocess부터 다시 시작 
Cord_bysT <- readRDS("/home/jaehyunchoi/project/Byst_collaboration/analysis/Cord_byst_ver2.rds")
gene_symbols <- rownames(Cord_bysT)
rowData(cds)$gene_short_name <- gene_symbols
cds <- as.cell_data_set(Cord_bysT, )
cds <- preprocess_cds(cds, num_dim = 100)
#pca check
pdf("/home/jaehyunchoi/project/Byst_collaboration/analysis/monocle3_plot_cds_pca.pdf", width = 8, height = 6)
plot_pc_variance_explained(cds)
dev.off()
#
cds <- reduce_dimension(cds)
pdf("/home/jaehyunchoi/project/Byst_collaboration/analysis/monocle3_plot_cds_reduc.pdf", width = 8, height = 6)
plot_cells(cds)
dev.off()

colData(cds)$Celltype <- Cord_bysT$celltype
pdf("/home/jaehyunchoi/project/Byst_collaboration/analysis/monocle3_plot_cds_celltype.pdf", width = 8, height = 6)
plot_cells(cds, color_cells_by="Celltype")
dev.off()
pdf("/home/jaehyunchoi/project/Byst_collaboration/analysis/monocle3_plot_cds_geneexpr.pdf", width = 8, height = 6)
plot_cells(cds, genes=c("Areg", "Il13", "Il5", "Pdcd1","Foxp3"))
dev.off()